---
title: "Differential Expression Analysis with DESeq2"
subtitle: "GSE190752 bulk RNA-seq"
author: "Seyedeh Zahra Mousavi"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
    html_preview: false
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
if (.Platform$OS.type == "windows") {
  windowsFonts(Arial = windowsFont("Arial"))
}
```

# Overview

This notebook performs differential expression analysis on bulk RNA-seq count data using [DESeq2](https://bioconductor.org/packages/DESeq2/).

**Expected input files (same folder as this `.Rmd`):**

| File | Description |
|------|-------------|
| `counts.txt` | Gene count matrix (genes as rows, samples as columns) |
| `Pheno.txt` | Sample metadata / phenotype table (samples as rows) |

> **Note:** Change the working directory and file names if your paths differ. The original script used a Windows path (`C:/Users/...`). For GitHub and reproducibility, keep the data files next to the notebook or use a relative path.

---

# 1. Install and load packages

Run the installation block **once**. After that, only load the libraries.

```{r install-packages, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("DESeq2")
install.packages(c("ggplot2", "pheatmap"))
```

```{r libraries}
library(DESeq2)
library(ggplot2)
library(pheatmap)

theme_set(
  theme_bw(base_family = "Arial") +
    theme(text = element_text(family = "Arial"))
)
```

---

# 2. Working directory and input data

```{r paths}
# Folder that contains count.txt and Pheno.txt
data_dir <- "C:/Users/Z/Desktop/GSE190752/"
if (dir.exists(data_dir)) {
  setwd(data_dir)
}

count_file <- "counts.txt"
pheno_file <- "Pheno.txt"

cat("Working directory:\n", getwd(), "\n\n")
cat("Files in this folder:\n")
print(list.files())

if (!file.exists(count_file)) {
  stop("Cannot find ", count_file, " in ", getwd(),
       ". Put the count file in this folder or change count_file / data_dir.")
}
if (!file.exists(pheno_file)) {
  stop("Cannot find ", pheno_file, " in ", getwd(),
       ". Put the phenotype file in this folder or change pheno_file / data_dir.")
}
```

## Count matrix

```{r read-counts}
Counts <- read.delim(count_file, sep = "\t", row.names = 1)
colnames(Counts)
dim(Counts)
```

## Filter lowly expressed genes

Keep genes with a total count of at least 10 across all samples.

```{r filter-counts}
keep <- rowSums(Counts) >= 10
Counts <- Counts[keep, ]
dim(Counts)
```

## Phenotype / sample information

```{r read-pheno}
info <- read.delim(pheno_file, header = TRUE, row.names = 1)
rownames(info)
colnames(info)
```

## Align sample order

DESeq2 requires that columns of the count matrix match the rows of `colData`.

```{r order-samples}
stopifnot(all(rownames(info) %in% colnames(Counts)))
Counts_Ordered <- Counts[, rownames(info)]
identical(colnames(Counts_Ordered), rownames(info))
```

---

# 3. Build the DESeq2 object

The design formula uses the `status` column in the phenotype table.  
Change `~ status` if your grouping variable has another name.

```{r make-dds}
dds <- DESeqDataSetFromMatrix(
  countData = Counts_Ordered,
  colData   = info,
  design    = ~ status
)
dds
```

Set the reference level for the contrast (here: `"Normal"`).

```{r relevel}
# Column in colData is "status" (from the design formula).
# The original script assigned to dds$condition; we keep status consistent.
dds$status <- relevel(factor(dds$status), ref = "Normal")
```

---

# 4. Run DESeq2

```{r deseq}
dds <- DESeq(dds)
```

---

# 5. Results

```{r results}
res <- results(dds)
res
summary(res)
```

## Export

```{r write-output}
write.csv(as.data.frame(res), "RES_DESEQ2_Comparison2.csv")
```

---

# 6. Plots

```{r vst}
vsd <- vst(dds, blind = FALSE)
```

## PCA

```{r pca, fig.cap="PCA of VST-transformed counts"}
plotPCA(vsd, intgroup = "status") +
  theme_bw(base_family = "Arial") +
  theme(text = element_text(family = "Arial"))
```

## Sample heatmap

```{r sample-heatmap, fig.cap="Sample-to-sample distance heatmap"}
sample_dists <- dist(t(assay(vsd)))
sample_dist_matrix <- as.matrix(sample_dists)
pheatmap(
  sample_dist_matrix,
  clustering_distance_rows = sample_dists,
  clustering_distance_cols = sample_dists,
  annotation_col = info["status"],
  main = "Sample distances",
  fontsize = 10,
  fontfamily = "Arial"
)
```

## Boxplot after normalization

```{r boxplot-normalized, fig.cap="Normalized counts per sample"}
norm_counts <- counts(dds, normalized = TRUE)
status_levels <- unique(as.character(info[colnames(norm_counts), "status"]))
status_colors <- setNames(
  c("#4C78A8", "#F58518", "#54A24B", "#E45756", "#72B7B2")[seq_along(status_levels)],
  status_levels
)
box_cols <- status_colors[as.character(info[colnames(norm_counts), "status"])]

par(family = "Arial", mar = c(8, 5, 4, 2) + 0.1)
boxplot(
  log2(norm_counts + 1),
  las = 2,
  outline = FALSE,
  col = box_cols,
  border = "grey20",
  main = "Normalized counts",
  ylab = "log2(normalized count + 1)",
  xlab = ""
)
legend(
  "topright",
  legend = names(status_colors),
  fill = status_colors,
  bty = "n",
  cex = 0.9
)
```

## Volcano plot

DEGs: FDR < 0.01 and |log2 fold change| > 1.

```{r volcano, fig.cap="Volcano plot"}
res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)

padj_cutoff <- 0.01
lfc_cutoff  <- 1

res_df$sig <- "NS"
res_df$sig[!is.na(res_df$padj) & res_df$padj < padj_cutoff & res_df$log2FoldChange >  lfc_cutoff] <- "Up"
res_df$sig[!is.na(res_df$padj) & res_df$padj < padj_cutoff & res_df$log2FoldChange < -lfc_cutoff] <- "Down"
res_df$sig <- factor(res_df$sig, levels = c("Up", "Down", "NS"))

ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
  geom_point(alpha = 0.6, size = 1.2, na.rm = TRUE) +
  scale_color_manual(values = c(Up = "#D73027", Down = "#4575B4", NS = "grey70")) +
  geom_vline(xintercept = c(-lfc_cutoff, lfc_cutoff), linetype = "dashed") +
  geom_hline(yintercept = -log10(padj_cutoff), linetype = "dashed") +
  theme_bw(base_family = "Arial") +
  theme(text = element_text(family = "Arial")) +
  labs(
    title = "Volcano plot (FDR < 0.01, |log2FC| > 1)",
    x = "log2 fold change",
    y = "-log10 FDR",
    color = "Significance"
  )
```

---

# Session info

```{r session-info}
sessionInfo()
```
