---
title: "Two-color Agilent microarray analysis with limma"
subtitle: "GSE145467"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 3
    html_preview: false
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6
)
if (.Platform$OS.type == "windows") {
  windowsFonts(Arial = windowsFont("Arial"))
}
```

# Overview

This notebook analyzes two-color Agilent arrays with [limma](https://bioconductor.org/packages/limma/).

Expected files next to this notebook (or under the paths set below):

| File / folder | Description |
|---------------|-------------|
| `pheno.txt` | limma targets file (`Cy5`, `Cy3`, file names) |
| `spot.txt` | Spot-type file for control probes |
| `GSE145467_RAW/` | Raw Agilent files listed in the targets file |

Change `data_dir` if your files live somewhere else.

# 1. Packages

```{r install-packages, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("limma")
install.packages("ggplot2")
```

```{r libraries}
library(limma)
library(ggplot2)

theme_set(
  theme_bw(base_size = 13, base_family = "Arial") +
    theme(
      text = element_text(family = "Arial", color = "grey15"),
      plot.title = element_text(face = "bold", hjust = 0.5),
      panel.grid.minor = element_blank()
    )
)

set_base_plot <- function() {
  par(
    family = "Arial",
    las = 2,
    cex.main = 1.1,
    cex.lab = 1,
    cex.axis = 0.85,
    bty = "l",
    mar = c(8, 5, 4, 2) + 0.1
  )
}

group_palette <- c("#4C78A8", "#F58518", "#54A24B", "#E45756", "#72B7B2", "#B279A2")
```

# 2. Paths and phenotype

```{r paths}
# Original analysis folder (change if needed)
data_dir <- "C:/Users/Z/Desktop/Microarray_NOA"

if (dir.exists(data_dir)) {
  setwd(data_dir)
}

targets_file <- "pheno.txt"
spot_file    <- "spot.txt"
raw_dir      <- "GSE145467_RAW"

cat("Working directory:\n", getwd(), "\n\n")
print(list.files())

if (!file.exists(targets_file)) {
  stop("Cannot find ", targets_file, " in ", getwd())
}
if (!file.exists(spot_file)) {
  stop("Cannot find ", spot_file, " in ", getwd())
}
if (!dir.exists(raw_dir)) {
  stop("Cannot find folder ", raw_dir, " in ", getwd())
}
```

```{r phenodata}
phenodata <- readTargets(targets_file)

# Clean Cy3 / Cy5 labels (encoding and non-alphanumeric characters)
phenodata$Cy5 <- iconv(phenodata$Cy5, to = "UTF-8", sub = "")
phenodata$Cy3 <- iconv(phenodata$Cy3, to = "UTF-8", sub = "")
phenodata$Cy5 <- gsub("[^a-zA-Z0-9]", "", phenodata$Cy5)
phenodata$Cy3 <- gsub("[^a-zA-Z0-9]", "", phenodata$Cy3)

head(phenodata)
```

# 3. Read arrays

```{r read-raw}
data <- read.maimages(
  files = phenodata,
  source = "agilent",
  path = raw_dir,
  green.only = FALSE
)
```

```{r plot-md-raw, fig.cap="MA plot of raw arrays"}
set_base_plot()
plotMD(data)
```

```{r spot-status}
spot <- readSpotTypes(file = spot_file)
data$genes$Status <- controlStatus(types = spot, genes = data)
set_base_plot()
plotMD(data)
```

# 4. Spot filter and weighted import

Keep a spot if the foreground / background ratio is > 4.5 in Cy5 or Cy3.

```{r filter-fun}
filter <- function(x) {
  R <- x[, "rMedianSignal"] / x[, "rBGMedianSignal"] > 4.5
  G <- x[, "gMedianSignal"] / x[, "gBGMedianSignal"] > 4.5
  as.numeric(R | G)
}
```

```{r read-weighted}
data2 <- read.maimages(
  files = phenodata,
  source = "agilent",
  path = raw_dir,
  green.only = FALSE,
  wt.fun = filter
)
data2$genes$Status <- controlStatus(types = spot, genes = data2)
```

```{r plot-md-filtered, fig.cap="MA plot after spot weighting"}
set_base_plot()
plotMD(data2)
```

```{r raw-boxplots, fig.cap="Raw foreground and background intensities", fig.width=9, fig.height=16}
stack_log <- function(mat, panel) {
  df <- stack(as.data.frame(log2(mat)))
  colnames(df) <- c("intensity", "sample")
  df$panel <- panel
  df
}

raw_long <- rbind(
  stack_log(data2$G, "Cy3 foreground"),
  stack_log(data2$Gb, "Cy3 background"),
  stack_log(data2$R, "Cy5 foreground"),
  stack_log(data2$Rb, "Cy5 background")
)

raw_long$panel <- factor(
  raw_long$panel,
  levels = c("Cy3 foreground", "Cy3 background", "Cy5 foreground", "Cy5 background")
)
raw_long$sample <- factor(raw_long$sample, levels = rev(colnames(data2$G)))

fill_cols <- c(
  "Cy3 foreground" = "#54A24B",
  "Cy3 background" = "#B8E0A8",
  "Cy5 foreground" = "#E45756",
  "Cy5 background" = "#F4A3A3"
)

ggplot(raw_long, aes(x = sample, y = intensity, fill = panel)) +
  geom_boxplot(color = "grey20", outlier.size = 0.4, show.legend = FALSE) +
  scale_fill_manual(values = fill_cols) +
  coord_flip() +
  facet_wrap(~ panel, ncol = 1, scales = "free_x") +
  labs(title = "Raw intensities", x = NULL, y = "log2 intensity") +
  theme(
    axis.text.y = element_text(family = "Arial", size = 7),
    strip.text = element_text(family = "Arial", face = "bold")
  )
```

# 5. Background correction and normalization

```{r normalize}
background_corrected <- backgroundCorrect(
  RG = data2,
  method = "normexp",
  offset = 15
)

MA <- normalizeWithinArrays(background_corrected, method = "loess")
norm_data <- normalizeBetweenArrays(object = MA, method = "quantile")
```

```{r plot-densities, fig.cap="Densities after normalization"}
set_base_plot()
plotDensities(norm_data)
```

```{r plot-m-boxplot, fig.cap="Normalized M-values", fig.width=9, fig.height=8}
M_long <- stack(as.data.frame(norm_data$M))
colnames(M_long) <- c("M", "sample")
M_long$sample <- factor(M_long$sample, levels = rev(colnames(norm_data$M)))

ggplot(M_long, aes(x = sample, y = M)) +
  geom_boxplot(fill = "#72B7B2", color = "grey20", outlier.size = 0.6) +
  coord_flip() +
  labs(
    title = "Normalized M-values",
    x = NULL,
    y = "M (log2 Cy5/Cy3)"
  ) +
  theme(
    axis.text.y = element_text(family = "Arial", size = 8),
    axis.text.x = element_text(family = "Arial", size = 10)
  )
```

# 6. Average replicate probes

```{r avereps}
ave <- avereps(x = norm_data, ID = norm_data$genes$ProbeName)
```

# 7. Linear model and contrast

Reference group in the targets file: `Control`.  
The contrast of interest is `impaired - normal`.

```{r model}
design <- modelMatrix(targets = phenodata, ref = "Control")
fit <- lmFit(object = ave, design = design)

contrast <- makeContrasts(
  contrasts = c("impaired", "normal", "impaired-normal"),
  levels = design
)

fit2 <- contrasts.fit(fit = fit, contrasts = contrast)
fit3 <- eBayes(fit2)
```

```{r toptable}
tt <- topTable(
  fit = fit3,
  number = Inf,
  adjust.method = "fdr",
  coef = "impaired-normal",
  sort.by = "logFC"
)

sig <- subset(tt, adj.P.Val < 0.05)
write.table(sig, "DEG_impaired_vs_normal.txt", quote = FALSE, sep = "\t")

up <- subset(sig, logFC > 1)
up <- subset(up, select = intersect(c("GeneName", "adj.P.Val", "logFC"), colnames(up)))
write.table(up, "DEG_up_logFC_gt1.txt", quote = FALSE, sep = "\t")

head(sig)
nrow(sig)
nrow(up)
```

# 8. PCA

```{r sample-groups}
samp_groups <- factor(phenodata$Cy5)
samp_cols <- group_palette[seq_along(levels(samp_groups))]
names(samp_cols) <- levels(samp_groups)
```

```{r pca, fig.cap="PCA of averaged M-values"}
M <- ave$M
keep <- complete.cases(M)
pca <- prcomp(t(M[keep, , drop = FALSE]), scale. = TRUE)
pct <- 100 * pca$sdev^2 / sum(pca$sdev^2)

pca_df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  group = samp_groups[match(rownames(pca$x), colnames(M))]
)

ggplot(pca_df, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 3.5, alpha = 0.9) +
  geom_text(aes(label = group), family = "Arial", size = 3.2, vjust = -1, show.legend = FALSE) +
  scale_color_manual(values = samp_cols) +
  labs(
    title = "PCA of averaged M-values",
    x = paste0("PC1 (", round(pct[1], 1), "%)"),
    y = paste0("PC2 (", round(pct[2], 1), "%)"),
    color = "Group"
  )
```

# Session info

```{r session-info}
sessionInfo()
```
